---
title: 'Reproducible Research: Peer Assessment 1'
author: "Anh"
date: "8/20/2021"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Loading and preprocessing the data
Load necessary libraries

```{r echo = TRUE}
library(ggplot2)
library(dplyr)
library(lattice)
```
Read the file
```{r echo = TRUE}
activity = read.csv("activity.csv")
```
Process datetime column of the data
```{r echo = TRUE}
activity$fixed_date = as.Date(activity$date)
activity$week_day = weekdays(activity$fixed_date)
```
## What is mean total number of steps taken per day?
Calculate the total number of steps taken per date (fixed_date)
```{r echo = TRUE}
total_steps = activity %>%
                group_by(fixed_date) %>%
                summarise(sum(steps))
```
Create a histogram of the total number of steps per day
```{r echo = TRUE}
hist(total_steps$`sum(steps)`, breaks = 5, xlab = "Steps", main = 
       "Total steps per day histogram", col = "blue")
```

Calculate the mean and median of the total number of steps per day
```{r echo = TRUE}
(mean(total_steps$`sum(steps)`, na.rm = TRUE))
(median(total_steps$`sum(steps)`, na.rm = TRUE))
```
## What is the average daily activity pattern?
Create a dataset of steps in 5-minute interval and take their average
```{r echo - TRUE}
intervals <- aggregate(steps ~ interval, activity, mean)
```
Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and 
the average number of steps taken, averaged across all days (y-axis)
```{r echo = TRUE}
ggplot(intervals, aes(x = interval, y = steps), xlab = "Intervals", ylab = 
         "Average number of steps") + 
  geom_line() +
  ggtitle("Average Number of Steps per Interval")
```

Which 5-minute interval, on average across all the days in the dataset, contains
the maximum number of steps?
```{r echo = TRUE}
max_steps = max(intervals$steps)
(intervals[intervals$steps == max_steps, ])
```
## Imputing missing values
Calculate the total number of missing values in the dataset
```{r echo = TRUE}
nrow(activity[is.na(activity$steps), ])
```
For this exercise, I have chosen to impute NA values by the mean of the 5-minute
intervals of the weekdays.
```{r echo = TRUE}
activity_wo_na = activity %>%
  group_by(week_day, interval) %>%
  mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))
```
Aggregate the total number of steps per day in the new dataset with NAs imputed
```{r echo = TRUE}
total_steps_wo_na = activity_wo_na %>%
     group_by(fixed_date) %>%
     summarise(sum(steps))
```

```{r echp = TRUE}
(mean(total_steps_wo_na$`sum(steps)`))
(median(total_steps_wo_na$`sum(steps)`))
```
Create a histogram of total number of steps per day, categorised by two datasets
(one with NAs and one with NAs imputed)
```{r echo = TRUE}
hist(total_steps_wo_na$`sum(steps)`, breaks = 5, xlab = "Steps", 
     main = "Total Steps per Day with NAs Fixed",col = "Black")
hist(total_steps$`sum(steps)`, breaks = 5, xlab = "Steps", 
     main = "Total Steps per Day with NAs Fixed", col = "Grey", add = TRUE)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=  c("black", "grey"))
```

## Are there differences in activity patterns between weekdays and weekends?
Create a new column to determine if the week_day is weekday or weekend
```{r echo = TRUE}
activity_wo_na$week_day_or_weekend = ifelse(activity_wo_na$week_day %in%
                                              c("Saturday", "Sunday"),
                                            "weekend", "weekday")
```
Summarise the number of steps by intervals and weekday category (weekdays or 
weekend)
```{r echo = TRUE}
by_intervals_weekday = activity_wo_na %>%
                        group_by(interval, week_day_or_weekend) %>%
                        summarise(avg = mean(steps))
```
Plot
```{r echo = TRUE}
xyplot(avg ~ interval|week_day_or_weekend, data = by_intervals_weekday, 
       type = "l", layout = c(1,2),
       main = "Average Steps per Interval Based on Type of Day", 
       ylab = "Average Number of Steps", xlab = "Interval")
```